---
title: "State DPT3 Coverage Verification"
author: "Taiwo Adegbola"
format: html
editor: visual
date: "`r format(Sys.Date(), '%d %B %Y')`"
---

## 1. Background

Diphtheria, pertussis, and tetanus (DPT) remain vaccine-preventable diseases of major public health importance, particularly among children under five years of age. Achieving and sustaining high coverage with the third dose of the DPT-containing vaccine (DPT3) is widely used as a key indicator of the performance of routine immunization programmes.

The World Health Organization (WHO) recommends that countries achieve **at least 90% national DPT3 coverage** to provide adequate population-level protection and reduce the risk of outbreaks of vaccine-preventable diseases. Failure to attain this benchmark may indicate gaps in vaccine delivery, inequitable access to immunization services, or weaknesses in health system performance.

Despite ongoing efforts to strengthen routine immunization in Nigeria, substantial variation in DPT3 coverage persists across states. Understanding whether the observed average state-level coverage differs from the WHO-recommended target is important for evaluating programme performance and informing national immunization strategies.

This study uses state-level DPT3 coverage data to assess whether Nigeria's mean DPT3 coverage is statistically consistent with the WHO target of 90%.

## 2. Epidemiologic Question

**Is the average DPT3 vaccination coverage across Nigerian states statistically different from the WHO-recommended national target of 90%?**

## 3. Objective

To evaluate whether the mean DPT3 vaccination coverage across Nigerian states is statistically different from the WHO-recommended national coverage target of 90%.

## 4. Statistical method

**Method:** One-Sample t-test

**Rationale:** A one-sample t-test is appropriate because the analysis compares the mean DPT3 coverage from a single sample of Nigerian states with a fixed reference value (the WHO target of 90%).

## 5. Reference value

The reference value for this analysis is the **World Health Organization (WHO) recommended DPT3 immunization coverage target of 90%**, which is considered the minimum coverage required to achieve population-level (herd) immunity against diphtheria, pertussis, and tetanus. This benchmark serves as the hypothesized population mean against which the observed state-level DPT3 coverage is statistically evaluated.

## 6. Data source

State-level DPT3 vaccination coverage data were compiled from published secondary sources.

Three state-specific coverage estimates—**Ebonyi (98.7%)**, **Sokoto (11.5%)**, and the **Federal Capital Territory (79.6%)**—were obtained directly from **Olulaja & Jaber (2026, *Public Health Challenges*)** and independently corroborated using **Tribune Online**.

The published report also confirmed that the following seven states achieved **≥80%** DPT3 coverage:

- Ebonyi, Enugu, Imo, Ekiti, Lagos, Edo, Osun

and that the following seven states recorded **\<50%** coverage:

- Sokoto, Zamfara, Borno, Bauchi, Gombe, Niger, Katsina

Because complete state-specific coverage estimates for the remaining **23 states** were not publicly available, their values were **synthetically generated** within the reported coverage bands while preserving the published national mean of approximately **56%**. The resulting dataset therefore satisfies all published constraints but does **not** represent official surveillance records.

Actual district (LGA)-level DPT3 coverage data are maintained within the Nigerian **District Health Information System 2 (DHIS2)** and are not publicly accessible.

Accordingly, this dataset should be regarded as a **synthetic dataset calibrated to verified published statistics** and is intended solely to demonstrate epidemiologic and statistical methodology. Findings should therefore be interpreted as **illustrative of analytical methods rather than official programmatic estimates or surveillance conclusions**.

```{r libraries}

library(tidyverse)
library(janitor)
library(rio)
library(broom)
library(gt)
library(here)
```

```{r load-data}

# State-level DPT3 vaccination coverage data -from published secondary sources

dpt3_coverage <- import("dpt3_state_coverage_v2.csv") |> janitor::clean_names()
dpt3_coverage
```

```{r WHO-target}

dpt3_coverage <- dpt3_coverage |>
  mutate(
    who_target = if_else(
      dpt3_coverage_percent >= 90,
      "Met Target",
      "Below Target"
    )
  )
```

```{r Descriptive Statistics}


dpt3_summary <- dpt3_coverage |>
  summarise(
    Number_of_States = n(),
    Mean_Coverage = mean(dpt3_coverage_percent),
    Standard_Deviation = sd(dpt3_coverage_percent),
    Minimum_Coverage = min(dpt3_coverage_percent),
    Maximum_Coverage = max(dpt3_coverage_percent)
  )

dpt3_summary


ggplot(
  dpt3_coverage,
  aes(
    x = reorder(state, dpt3_coverage_percent),
    y = dpt3_coverage_percent,
    fill = who_target
  )
) +
  geom_col(width = 0.8) +
  
  geom_hline(
    yintercept = 90,
    colour = "red3",
    linewidth = 1.2,
    linetype = "dashed"
  ) +
  
  coord_flip() +
  
  scale_fill_manual(
    values = c(
      "Met Target" = "#238B45",
      "Below Target" = "#8FBBD9"
    )
  ) +
  
  labs(
    title = "State-level DPT3 Vaccination Coverage",
    subtitle = "Comparison with WHO Recommended Coverage Target (90%)",
    x = NULL,
    y = "DPT3 Coverage (%)",
    fill = NULL,
    caption = paste(
      "Source: Synthetic dataset calibrated to published DPT3 coverage statistics",
      "(Olulaja & Jaber, 2026).",
      "Red dashed line indicates the WHO recommended target (90%)."
    )
  ) +
  
  theme_minimal(base_size = 13) +
  
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 12),
    plot.caption = element_text(size = 9, colour = "grey40"),
    panel.grid.minor = element_blank(),
    legend.position = "right"
  )
  

# Create a North/South region variable from your existing Zone column

dpt3_coverage <- dpt3_coverage %>%
  mutate(region = case_when(
    zone %in% c("North Central", "North East", "North West") ~ "North",
    zone %in% c("South East", "South South", "South West") ~ "South"
  ))

# Labeled, ordered bar chart
library(ggplot2)

ggplot(dpt3_coverage, aes(x = reorder(state, dpt3_coverage_percent),
                          y = dpt3_coverage_percent, fill = region)) +
  geom_col() +
  geom_hline(yintercept = 90, linetype = "dashed", color = "red") +
  coord_flip() +
  labs(
    title = "DPT3 Coverage by State, Ordered Low to High",
    subtitle = "Red dashed line = WHO 90% threshold",
    x = "State", y = "DPT3 Coverage (%)"
  ) +
  theme_minimal()


```

```{r Checking the one-sample t-test assumptions}

# Shapiro-Wilk Test

shapiro.test(dpt3_coverage$dpt3_coverage_percent)

# Q-Q Plot
ggplot(
  dpt3_coverage,
  aes(sample = dpt3_coverage_percent)
) +
  stat_qq() +
  stat_qq_line(colour = "red") +
  labs(
    title = "Normal Q-Q Plot of DPT3 Coverage"
  ) +
  theme_minimal()

# Histogram

ggplot(
  dpt3_coverage,
  aes(x = dpt3_coverage_percent)
) +
  geom_histogram(
    aes(y = after_stat(density)),
    bins = 10,
    fill = "#2C7FB8",
    colour = "white",
    alpha = 0.7
  ) +
  
  geom_density(
    colour = "red",
    linewidth = 1.2
  ) +
  
  labs(
    title = "Distribution of State-level DPT3 Coverage",
    subtitle = "Histogram with kernel density estimate",
    x = "DPT3 Coverage (%)",
    y = "Density"
  ) +
  
  theme_minimal(base_size = 13)

```

```{r- Hypotheses}

Null hypothesis (H₀):

The mean DPT3 vaccination coverage across Nigerian states is equal to the WHO-recommended target of 90%.

                               g H1: μ = 90%

Alternative hypothesis (H₁):

The mean DPT3 vaccination coverage across Nigerian states is different from the WHO-recommended target of 90%.

                              H1: μ ≠ 90%
```

```{r t-test}
  t_result <- t.test(dpt3_coverage$dpt3_coverage_percent, mu = 90, alternative = "two.sided")
  t_result

```

```{r tidy-output}

broom::tidy(t_result) %>%
  select(estimate, statistic, p.value, conf.low, conf.high) %>%
  mutate(across(where(is.numeric), ~round(.x, 2)))
```

## 7. Result

A one-sample t-test was conducted to determine whether the mean DPT3 vaccination coverage across Nigerian states differed significantly from the WHO-recommended target of [<strong>90%</strong>]{.text-info}.

The analysis showed that the mean DPT3 coverage ([<strong>55.1%</strong>]{.text-success}) was significantly lower than the WHO target of [<strong>90%</strong>]{.text-info} ([<strong>t(36) = −8.09</strong>]{.text-primary}, [<strong>p = 1.304 × 10⁻⁹</strong>]{.text-danger}).

The [<strong>95% confidence interval</strong>]{.text-danger} for the population mean DPT3 coverage ranged from [<strong>46.35%</strong>]{.text-danger} to [<strong>63.85%</strong>]{.text-danger}, which does not include the WHO target of [<strong>90%</strong>]{.text-info}.

Therefore, the null hypothesis was rejected. There is sufficient statistical evidence to conclude that the mean DPT3 vaccination coverage across Nigerian states is significantly lower than the WHO-recommended coverage target of [<strong>90%</strong>]{.text-info}.

------------------------------------------------------------------------

## Interpretation

A one-sample t-test indicated that the mean DPT3 vaccination coverage across Nigerian states ([<strong>M = 55.1%</strong>]{.text-success}, SD = 26.25) was significantly lower than the WHO-recommended coverage threshold of [<strong>90%</strong>]{.text-info} ([<strong>t(36) = −8.09</strong>]{.text-primary}, [<strong>p \< .001</strong>]{.text-danger}, [<strong>95% CI: 46.35%–63.85%</strong>]{.text-danger}). The standardized effect size was [<strong>large (Cohen's d = −1.33)</strong>]{.text-success}, indicating that the observed shortfall from the WHO target was substantial rather than merely statistically significant.

### Statistical vs. Practical Significance

The large effect size demonstrates that the gap between the observed mean DPT3 coverage and the WHO-recommended target is not only statistically significant but also of considerable public health importance. On average, Nigerian states remain well below the immunization coverage required to achieve and sustain population-level immunity against diphtheria, pertussis, and tetanus.

[<strong>However, important geographic heterogeneity exists.</strong>]{.text-warning} While a few states (e.g., Ebonyi) have already achieved or exceeded the WHO target, several northern states remain far below the recommended threshold. Consequently, the national shortfall reflects marked inequalities in routine immunization performance rather than uniformly poor coverage across all states. These findings underscore the need for targeted immunization strengthening strategies in low-performing states while sustaining gains in states that have already achieved high coverage.

```{The DPT3 coverage data appear bimodal, reflecting two distinct regional clusters rather than a single approximately normal population.}

Null hypothesis (H₀):
  
  There is no difference in the distribution of DPT3 coverage between the North and South (location shift = 0).

Alternative hypothesis (H₁):
  
  The distributions of DPT3 coverage differ between the North and South (location shift ≠ 0).

```

```{r wilcoxon test}

wilcox.test(
  dpt3_coverage_percent ~ region,
  data = dpt3_coverage
)
```
