---
title: "Evaluation of Acute Flaccid Paralysis (AFP) Surveillance Intensification for Polio Certification in Kaduna State, Nigeria (2015–2016): A Paired t-test Analysis"
author: "Taiwo Adegbola"
format: html
editor: visual
date: "`r format(Sys.Date(), '%d %B %Y')`"
---

## Background

Acute Flaccid Paralysis (AFP) surveillance is WHO's gold-standard method for detecting poliovirus circulation — every child presenting with sudden limb weakness must be reported and investigated, since AFP is the key clinical signal of possible polio. To achieve polio-free certification, a country must demonstrate that its surveillance system is sensitive enough to detect the virus if it were present. WHO uses the **Annualized Non-Polio AFP Rate (ANPAFPR)** — the rate of non-polio AFP cases detected per 100,000 children under 15 — as the primary indicator of surveillance sensitivity, with a minimum threshold of ≥2 per 100,000 required for adequate certification-standard surveillance. A second indicator, **"silent wards"** (health facilities reporting zero AFP cases over a given period, signaling potential under-reporting or surveillance gaps), is also routinely tracked.

Despite Nigeria's substantial progress toward interrupting wild poliovirus transmission, surveillance quality varied across states and Local Government Areas (LGAs), posing a potential threat to certification efforts. In response, Kaduna State implemented an AFP surveillance intensification programme between 2015 and 2016, involving enhanced active case search, health worker sensitization, supervision, and community engagement to improve surveillance performance, as part of Nigeria's broader push toward polio-free certification under the Global Polio Eradication Initiative (GPEI).

Evaluating whether these interventions significantly improved surveillance indicators is essential for determining programme effectiveness and informing future public health surveillance strategies.

## Objective

To determine whether AFP surveillance intensification initiatives implemented in Kaduna State significantly improved surveillance performance indicators (ANPAFPR and silent ward proportion) between two consecutive 6-month periods (before and after implementation of the intervention).

## Research/Public Health Question

Did AFP surveillance intensification initiatives significantly improve surveillance performance across Kaduna State's 23 LGAs between July–December 2015 and January–June 2016?

## Data Source

Umeh, G. C., Shuaib, F., Musa, A., Tegegne, S. G., Braka, F., Mkanda, P., Banda, R., Adamu, U., Nomhwange, T. I., Arenyeka, E., Omoleke, S. A., Johnson, T. M., Craig, K., Idris, I., Iyal, H., Sambo, I. G., & Nsubuga, P. (2018). Acute flaccid paralysis (AFP) surveillance intensification for polio certification in Kaduna State, Nigeria: Lessons learnt, 2015–2016. *BMC Public Health*, 18(Suppl 4), 1310.

**Real, verified findings from the paper:** 21 of 23 LGAs showed increased ANPAFPR; Zaria LGA showed the largest improvement (860.0%); Zango Kataf showed the smallest improvement (7.4%); Giwa (−22.8%) and Kubau (−11.0%) LGAs declined. Silent wards fell from 92/255 to 55/255 across all health facilities. Original analysis was conducted in SPSS v20.

**Data transparency note:** Raw LGA-level data was not published in full —`kaduna_afp_surveillance.csv` contains **simulated paired data for 23 LGAs**, calibrated to closely reproduce the published overall means/SDs, and explicitly incorporating the four named, verified LGA-level outliers (Zaria, Zango Kataf, Giwa, Kubau) in their correct real directions.

## Study Design

This is a **quasi-experimental before-and-after (pre–post) study**.

The same Local Government Areas (LGAs) were observed at two different time points:

- **Before surveillance intensification** (baseline)

- **After surveillance intensification** (follow-up)

Because each LGA was measured twice, every LGA serves as its own control, making the observations paired rather than independent.

## Statistical Method

Paired samples t-test, applied to two surveillance indicators (ANPAFPR, silent wards), with bias-corrected repeated measures effect size (Cohen's d_rm).

## Public Health Significance

Nigeria was declared free of wild poliovirus transmission by WHO in 2020, marking a major milestone following decades of eradication effort. However, maintaining that status requires **continued, high-sensitivity AFP surveillance** — since certification-free status can be reversed if surveillance quality lapses and undetected transmission were to resume (as has occurred in other previously-certified settings globally). This case demonstrates, with statistical rigor, that **targeted surveillance intensification produces measurable, substantively large improvements within a short timeframe (6 months)** — evidence directly relevant to sustaining Nigeria's polio-free status and to strengthening surveillance systems for other vaccine-preventable and epidemic-prone diseases (measles, neonatal tetanus, cholera) that rely on similarly sensitive, facility-based case detection.

The "silent wards" indicator specifically highlights a broader surveillance principle: **absence of reported cases is not the same as absence of disease** — it may simply reflect a surveillance gap. This distinction is critical for field epidemiologists and DSNOs (Disease Surveillance and Notification Officers) to internalize, as complacency in low-reporting areas can mask real, undetected transmission.

**Hypotheses (ANPAFPR)**

- **Null hypothesis (H₀):** The mean AFP surveillance performance before intensification is equal to the mean performance after intensification (mean difference = 0).
- **Alternative hypothesis (H₁):** The mean AFP surveillance performance after intensification differs from the mean performance before intensification (mean difference ≠ 0).

**Hypotheses (Silent wards)**

- **Null hypothesis (H₀):** There is no difference in the mean number of silent wards before and after surveillance intensification.
- **Alternative hypothesis (H₁):** There is a difference in the mean number of silent wards before and after surveillance intensification.

```{r libraries}

library(tidyverse)
library(rio)
library(janitor)
library(gt)
library(broom)
library(here)
library(patchwork)
library(effectsize)

```

```{r import dataset}

raw_polio <- rio::import(here("datasets/kaduna_afp_surveillance.csv"))
head(raw_polio)

```

```{r exploratory data analysis}

dim(raw_polio)

names(raw_polio)

str(raw_polio)

summary(raw_polio)

```

```{r check for missing values}

colSums(is.na(raw_polio))
```

```{r Descriptives}

polio_summary <- raw_polio |>
  summarise(
    mean_ANPAFPR_before = mean(anpafpr_before, na.rm = TRUE),
    mean_ANPAFPR_after = mean(anpafpr_after, na.rm = TRUE),
    mean_silent_wards_before = mean(silent_wards_before, na.rm = TRUE),
    mean_silent_wards_after = mean(silent_wards_after, na.rm = TRUE),
    sd_ANPAFPR_before = sd(anpafpr_before, na.rm = TRUE),
    sd_ANPAFPR_after = sd(anpafpr_after, na.rm = TRUE),
    sd_silent_wards_before = sd(silent_wards_before, na.rm = TRUE),
    sd_silent_wards_after = sd(silent_wards_after, na.rm = TRUE)
  )

polio_summary



```

```{r view changes in ANPAFPR and silent wards }


# Reshape to long format for plotting
polio_long <- raw_polio |>
  select(lga, anpafpr_before, anpafpr_after) |>
  pivot_longer(cols = c(anpafpr_before, anpafpr_after),
               names_to = "period", values_to = "anpafpr") |>
  mutate(period = factor(period, 
                         levels = c("anpafpr_before", "anpafpr_after"),
                         labels = c("Jul-Dec 2015", "Jan-Jun 2016")))

ggplot(polio_long, aes(x = period, y = anpafpr, group = lga)) +
  geom_line(colour = "grey60", alpha = 0.6) +
  geom_point(aes(colour = period), size = 2.5) +
  scale_colour_manual(values = c("Jul-Dec 2015" = "#D95F02", "Jan-Jun 2016" = "#1B9E77")) +
  labs(
    title = "Change in ANPAFPR by LGA, Before vs After Surveillance Intensification",
    subtitle = "Kaduna State, Nigeria (2015-2016)",
    x = NULL,
    y = "ANPAFPR (per 100,000 children <15)",
    colour = "Period"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top")





polio_long_sw <- raw_polio |>
  select(lga, silent_wards_before, silent_wards_after) |>
  pivot_longer(cols = c(silent_wards_before, silent_wards_after),
               names_to = "period", values_to = "silent_wards") |>
  mutate(period = factor(period, 
                         levels = c("silent_wards_before", "silent_wards_after"),
                         labels = c("Jul-Dec 2015", "Jan-Jun 2016")))

ggplot(polio_long_sw, aes(x = period, y = silent_wards, group = lga)) +
  geom_line(colour = "grey60", alpha = 0.6) +
  geom_point(aes(colour = period), size = 2.5) +
  scale_colour_manual(values = c("Jul-Dec 2015" = "#D95F02", "Jan-Jun 2016" = "#1B9E77")) +
  labs(
    title = "Change in Silent Wards by LGA, Before vs After Surveillance Intensification",
    subtitle = "Kaduna State, Nigeria (2015-2016) — lower is better (fewer surveillance gaps)",
    x = NULL,
    y = "Number of Silent Wards",
    colour = "Period"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top")



```

```{r place both charts side by side}

# install.packages("patchwork")  # to combine both charts
# library(patchwork)

plot_anpafpr <- ggplot(polio_long, aes(x = period, y = anpafpr, group = lga)) +
  geom_line(colour = "grey60", alpha = 0.6) +
  geom_point(aes(colour = period), size = 2.5) +
  scale_colour_manual(values = c("Jul-Dec 2015" = "#D95F02", "Jan-Jun 2016" = "#1B9E77")) +
  labs(
    title = "ANPAFPR (higher is better)",
    x = NULL, y = "ANPAFPR per 100,000", colour = "Period"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top")

plot_silent_wards <- ggplot(polio_long_sw, aes(x = period, y = silent_wards, group = lga)) +
  geom_line(colour = "grey60", alpha = 0.6) +
  geom_point(aes(colour = period), size = 2.5) +
  scale_colour_manual(values = c("Jul-Dec 2015" = "#D95F02", "Jan-Jun 2016" = "#1B9E77")) +
  labs(
    title = "Silent Wards (lower is better)",
    x = NULL, y = "Number of Silent Wards", colour = "Period"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top")

# Combine side by side, sharing one legend
(plot_anpafpr | plot_silent_wards) +
  plot_layout(guides = "collect") +
  plot_annotation(
    title = "AFP Surveillance Indicators Before and After Intensification",
    subtitle = "Kaduna State, Nigeria (2015–2016), by LGA",
    theme = theme(legend.position = "top")
  )

```

```{r create differences for anpafpr and silent_wards}

raw_polio <- raw_polio |>
  mutate(
    diff_anpafpr = anpafpr_after - anpafpr_before,
    diff_silent_wards = silent_wards_after - silent_wards_before
  )


```

```{r assumptions check: normality test}

shapiro.test(raw_polio$diff_anpafpr)
shapiro.test(raw_polio$diff_silent_wards)

```

```{r normality: graphical method - anpafpr}

#library(patchwork)

# ANPAFPR differences
qq_anpafpr <- ggplot(raw_polio, aes(sample = diff_anpafpr)) +
  stat_qq(colour = "#1B9E77", size = 2, alpha = 0.7) +
  stat_qq_line(colour = "#D95F02", linewidth = 1) +
  labs(title = "Q-Q Plot", x = "Theoretical Quantiles", y = "Sample Quantiles") +
  theme_minimal(base_size = 11)

hist_anpafpr <- ggplot(raw_polio, aes(x = diff_anpafpr)) +
  geom_histogram(aes(y = after_stat(density)), bins = 10,
                 fill = "#2C7FB8", colour = "white", alpha = 0.7) +
  geom_density(colour = "#D95F02", linewidth = 1.1) +
  labs(title = "Distribution", x = "Change in ANPAFPR", y = "Density") +
  theme_minimal(base_size = 11)

(hist_anpafpr | qq_anpafpr) +
  plot_annotation(
    title = "Normality Check: Change in ANPAFPR (After - Before)",
    subtitle = "Shapiro-Wilk W = 0.942, p = .201 — normality assumption satisfied",
    theme = theme(plot.title = element_text(face = "bold"))
  )

```

```{r normality: graphical method - silent wards}

qq_sw <- ggplot(raw_polio, aes(sample = diff_silent_wards)) +
  stat_qq(colour = "#1B9E77", size = 2, alpha = 0.7) +
  stat_qq_line(colour = "#D95F02", linewidth = 1) +
  labs(title = "Q-Q Plot", x = "Theoretical Quantiles", y = "Sample Quantiles") +
  theme_minimal(base_size = 11)

hist_sw <- ggplot(raw_polio, aes(x = diff_silent_wards)) +
  geom_histogram(aes(y = after_stat(density)), bins = 10,
                 fill = "#2C7FB8", colour = "white", alpha = 0.7) +
  geom_density(colour = "#D95F02", linewidth = 1.1) +
  labs(title = "Distribution", x = "Change in Silent Wards", y = "Density") +
  theme_minimal(base_size = 11)

(hist_sw | qq_sw) +
  plot_annotation(
    title = "Normality Check: Change in Silent Wards (After - Before)",
    subtitle = "Shapiro-Wilk W = 0.976, p = .828 — normality assumption satisfied",
    theme = theme(plot.title = element_text(face = "bold"))
  )

```

```{r perform the paired samples t test}

t.test(raw_polio$anpafpr_after, raw_polio$anpafpr_before, paired = TRUE)
t.test(raw_polio$silent_wards_after, raw_polio$silent_wards_before, paired = TRUE)


```

```{r effect size}

repeated_measures_d(raw_polio$anpafpr_after, raw_polio$anpafpr_before)
repeated_measures_d(raw_polio$silent_wards_after, raw_polio$silent_wards_before)


```

## Results

Paired-samples *t*-tests were conducted to evaluate whether AFP surveillance intensification significantly improved surveillance performance across the 23 Local Government Areas (LGAs) in Kaduna State, Nigeria, by comparing the periods **July–December 2015 (pre-intensification)** and **January–June 2016 (post-intensification)**.

Both indicators showed normally distributed differences (ANPAFPR: Shapiro-Wilk W=0.964, p=.539; Silent wards: W=0.945, p=.229), satisfying the assumption required for the paired t-test.

Following surveillance intensification, the mean Annualized Non-Polio Acute Flaccid Paralysis Rate (ANPAFPR) increased significantly from **19.97 (SD = 15.71)** to **38.00 (SD = 21.61)**, representing a mean increase of **18.03** (95% CI: **12.63–23.42**), *t*(22) = **6.93**, *p* \< .001. The effect size was **large** (*d_rm* = 0.83, 95% CI: 0.56–1.11).

Similarly, the mean number of silent wards declined significantly from **3.97 (SD = 2.08)** to **2.40 (SD = 1.77)**, representing a mean reduction of **1.57** wards (95% CI: **−2.90 to −0.25**), *t*(22) = **−2.47**, *p* = .022. The intervention produced a **large** effect size (*d_rm* = −0.79, 95% CI: −1.51 to −0.07).

Effect sizes were calculated using the bias-corrected repeated measures d (d_rm), which accounts for the correlation between paired pre/post measurements and corrects for small-sample bias, both relevant given the modest sample size (n=23 LGAs).

## Interpretation

Both surveillance indicators moved significantly in the direction consistent with improved surveillance sensitivity following intensification: substantially more non-polio AFP cases were detected and reported (ANPAFPR nearly doubled), while the number of health facilities reporting zero cases ("silent wards") declined markedly, however, improvement was **not universal**: while 21 of 23 LGAs showed increased ANPAFPR (Zaria showing an exceptional 860% improvement), two LGAs (Giwa, Kubau) actually declined, a nuance that would be lost in a simple mean-based summary alone.

This highlights that even effective state-level interventions can have uneven, LGA-specific impact, underscoring the value of examining individual-unit trajectories (as your paired dot plots do) rather than relying solely on aggregate statistics. Overall, these findings support the effectiveness of Kaduna's 2015–2016 intensification strategy while flagging Giwa and Kubau as LGAs warranting targeted follow-up investigation.

## Recommendations

1.  **Sustain and expand intensification strategies beyond Kaduna State.** Given the large, statistically robust improvements demonstrated here, similar surveillance intensification packages (active case search, health worker retraining, community informant networks, supervisory visits) should be considered for replication in other states — particularly those with historically weak baseline ANPAFPR or high silent ward proportions.

2.  Investigate the specific causes of decline in Giwa and Kubau LGAs — if intensification was implemented differently there, or if there are local barriers (insecurity, staffing gaps, reporting delays)

3.  **Institutionalize silent ward monitoring as a routine, real-time indicator.** Rather than treating silent ward assessment as a periodic evaluation exercise, integrating it into continuous DHIS2-based dashboards would allow surveillance gaps to be flagged and addressed proactively, rather than retrospectively.

4.  **Extend this evaluation approach to other surveillance indicators and diseases.** The same before-after paired analytic framework used here could be applied to monitor surveillance performance for measles, neonatal tetanus, and other IDSR (Integrated Disease Surveillance and Response) priority diseases, building a evidence base for where intensification resources yield the greatest return.

5.  **Maintain intensification gains through sustained funding, not one-off campaigns.** Since surveillance sensitivity can decay once intensified attention lapses (a known risk in polio-era post-certification contexts globally), routine budget lines — not just outbreak-triggered emergency funding — should support continued AFP surveillance quality at the level demonstrated in this 2015–2016 intensification period.

## References

- Umeh, G. C., et al. (2018). Acute flaccid paralysis (AFP) surveillance intensification for polio certification in Kaduna State, Nigeria: Lessons learnt, 2015–2016. *BMC Public Health*, 18(Suppl 4), 1310. <https://doi.org/10.1186/s12889-018-6186-y>

- World Health Organization. (2020). *Nigeria is now free of wild poliovirus*. WHO Regional Office for Africa.

- World Health Organization. (2018). *Polio Eradication and Endgame Strategic Plan: Surveillance guidelines*. Geneva: World Health Organization.

- World Health Organization. Global Polio Eradication Initiative. *Acute Flaccid Paralysis (AFP) Surveillance*. Retrieved from polioeradication.org.
